Drought-sensitive trees survive on deep water access in a tropical forest community

### 2020-07-15 manuscript.rmd
1. mortality analyses 2. discussion points 3. other smaller edits

Trait-based plant ecology aims to predict impacts of environmental change on community composition and function using relevant plant traits as currency. To reach this goal it is imperative to establish relationships between plant's environment, traits and demographic outcomes, but this remains to be established for trees' realised water environments and water-access through droughts. Huge efforts have been devoted to linking same-for-all-trees drought indices, above-ground traits and demographic rates. These have resulted in above-ground hydraulic traits such as proxies of hydraulic failure as suggested indicators of mortality risk. A recent meta-analysis found greater drought-induced mortality rates in species with high vulnerability to drought-induced embolism ($\Psi_{88,~Stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~Stem}$), but counter examples also exist (Hoffmann 2011; Paddock 2013), indicating whether plants actually reach critical thresholds of hydraulic failure depends on their realised drought intensity. A rare study that accounted for realised water availability for four tree species in a rare, centennial drought (Texas, USA) found that deep rooted species died less despite being more vulnerable to xylem embolism (high $\Psi_{88,~Stem}$) and having narrower safety-margins ($\Psi_{min} - \Psi_{88,~Stem}$). A species sourcing deep water showed lower hydraulic failure and lower mortality in an extreme drought in an Italian Karst system even though it had narrower hydrulic safety margin and high $\Psi_{88,~Stem}$ than other two species limited to shallow-water. Chitra-Tarak et al. (2018) proposed a general model to link realised water availability to demographic outcomes through droughts at tree community-scale. This study found that species accessing deep-water in a tropical dry forest study system died more in a rare, centennial drought compared to shallow-water species. Authors concluded that this was consistent with expectation of deep water usage as an aquisitive strategy relying on stable and reliable water resource at depth, which exhausted in the multi-annual drought resulting in reduced survival, compared to persistent species associated with shallow-water that was relatively more available via infiltration during this drought. However, species estimates of water-access modelled by this study were not corroborated with rooting/water-sourcing depths, or related to hydraulic traits.

@@

We test an overarching hypothesis that interspecific variation in effective rooting depths of trees aligns them on a conservative-acquisitive trade-off axis. Specifically, in this forest (i) dynamics of soil water potential at depth does not cross catastrophic levels even during multi-annual droughts because of high total annual precipitation to evapotranspiration ratio; (ii) trees with deeper effective rooting depths (ERD) show acquisitive hydraulic architecture: higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins; finally, (iii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure. To our knowledge this is the first study to demonstrate a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and mortality outcomes for a tropical forest community.

@@

```{r hydro-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap ="\\label{lab-hydro-calib}Model evaluation: Observations (red line) of Volumetric Water Content, Evapotranspiration and stream discharge aginst simulations (blue lines) from 100 best-fit parameter ensembles runs of ELM-FATES. This figure will be reformatted to compose individual figures into one."}
p1 <- cowplot::ggdraw() + cowplot::draw_image("figures/2019-10-14_5000/best-fits/swc_Obs_vs_model_daily_horizontal_with_steph_mean.jpeg", scale = 1)
p2 <- cowplot::ggdraw() + cowplot::draw_image("figures/2019-10-14_5000/best-fits/QRUNOFF_Obs_vs_model_monthly.jpeg", scale = 1)
p3 <- cowplot::ggdraw() +
  cowplot::draw_image("figures/2019-10-14_5000/best-fits/ET_Obs_vs_model_monthly.jpeg", scale = 1)
bottom_row <- cowplot::plot_grid(p2, p3, labels = c('B', 'C'), label_size = 12, ncol = 2)
cowplot::plot_grid(p1, bottom_row, labels = c('A', ''), label_size = 12, ncol = 1, rel_widths = c(2, 1))
```

## Model for Leaf Vulnerability Curves at community-scale

We measured ($K_{x,~Leaf}$ vs. $\Psi_{x,~Leaf}$) on adult trees of `r length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))` common species (see SI x). We fitted exponential, sigmoidal, logistic forms of equations to species-level data as prescribed by [@Guyot:2011] and based on AIC criteria chose exponential form (Fig. S\@ref(fig:obs-plc-curves)), described as

@@

```{r comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-plc-curves} Modelled standardized 'Vulnerability curves' ($K_{x,~Leaf}$ vs. $\\Psi_{x,~Leaf}$) at community-scale (", nrow(data.model.AB$sp)," species) based on species data on $WSG$ & $LMA$ and their relationships with vulnerability curve parameters measured on ", sp.soft.filled$AB.data.model," species. Each line represents a species, color coded by $WSG$.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB.jpeg"))
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_LMALAM_AVD_predicted_AB.jpeg"))
```

@@

With 
\
\begin{equation}
K_{Leaf,z,n} = Ae^{-B\Psi_{z,n}}
(\#eq:k-soil)
\end{equation}

@@

Growth Data: To minimise effect of light variation among trees and census intervals we selected only large trees (>= 30 cm DBH) and species whose maximum height was at least 30 m (hereafter, canopy species). We chose only those trees for which diameter (DBH) record was present across all size censuses (1990, 19905, 2000, 2005, 2010, 2015). We calculated individual tree ($i$) growth rates ($Gr_{i,t}$ cmyr^{-1}) for each of the five census intervals ($t$). As tree growth rates increase with size, to keep growth rates comparable across census intervals, we obtained residuals from a DBH model of tree growth based on a trees' DBH at the beginning of a census onterval ($GrResid_{i,t} = Gr_{i,t} - f(DBH_{i, t})$). The diameter model of growth was fitted based on all trees' data pooled together as there was insufficient species-specific data for large trees (SI ). We then standardised (centered and scaled) an individual tree time series ($GrResid_{i,t}$) and obtained species-level (*n* = 32) times-series of medians ($Growth_{t}$) that had data on at least three trees (median = xx).

@@

For each model (\@ref(eq:M2)- \@ref(eq:M4)) and species combination ($sp$), we obtained goodness-of-fit (*R*^2^) and Pearson's $\rho$ for all soil depths $z$, and for each sp, model we chose the $z$ with highest *R*^2^ as the Effective Rooting Depth (ERD); after retaining species with positive $\rho$ and *R*^2^ >= 0.1.

@@

Models with VPD and without Leaf Fraction (Eqs. \@ref(eq:M2)) had highest goodness-of fit (*R*^2^ = 0.51); followed by model with both VPD and Leaf Fraction (Eq. \@ref(eq:M4); Fig. S\@ref(fig:other-erd-mods)). Models with leaf fraction had only three species with sufficient, indvidual species-level goodness-of-fit (Eqs. \@ref(eq:M3 & eq:M4) versus seven for models without (Eqs. \@ref(eq:M1) & \@ref(eq:M2)). For rest of the analyses we therefore used species ERD estimates from model (\@ref(eq:M2)) as it was also a parsimonious model compared to \@ref(eq:M4) (Fig. \@ref(fig:erd-iso)).

```{r erd-iso, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-iso}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE; m) versus dry season stable isotopic concentration $\\delta$^2^H (mean $\\pm$ (add error); ‰) for ", length(unique(erd.data$sp))," canopy species from Barro Colorado Island (data from [@Meinzer:1999]")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_phenology_Meinzer_gr.Psi.VPD.jpeg"))
```

@@

We calculated mortality rates for large trees (>=10 cm DBH) of xx canopy species for each of the seven census interval ($t$) in the 33 year record (1985-2015; $Mort_{t}$) and a species mean mortality rate across those intervals ($Mort_{mean} = \frac{1}{t}\sum_{t=1}^{6} {Mort_{t}}$). To test whether $ERD$ plays a role in alleviating mortality risk, we regressed species $ERD$ against species ($Mort_{mean}$). To assess whether this may be due to deeper $ERD$ species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{x, ERD}$) were below species $\Psi_{80,Leaf}$.

@@

\
Species with deeper $ERD$ showed lower mean mortality rates across the 33 year record (1982-2015; *R*^2^ = 0.13, *P* = 0.029, *n* = xx; Figure \@ref(fig:erd-mort)). For five out of seven census intervals, species with shallower $ERD$ had significantly higher mortality anomalies ($Mort_{int}$ - $Mort{mean}$ than deeper ERD species; Figure (\@ref(fig:erd-mort-anom)). Analysis of time spent under catastrophic soil drought showed that the proportion of days species $\Psi_{x,~ERD}$ was more negative than species $\Psi_{80, Leaf}$ decreased with species ERD (Figure \@ref(fig:erd-time)). Species with ERD deeper than 1 m did not cross species $\Psi_{80, Leaf}$ at all for the entire 25 year period for which soil water potential $\Psi_{x,~z}$ was modelled (1990-2015).

```{r erd-mort, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort}Mortality rate (mean $\\pm$ (add error)) versus Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) for xx canopy species found on Barro Colorado Island. *R*^2^ and *P*-values are given in panel insets."}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_rate_by rdi.gr.jpeg"))
```

```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{x, z}$ was below $\\Psi_{80,Leaf}$ over 1990-2015, where z is the soil depth matching species ERD--for xx canopy species found on Barro Colorado Island. EAch circle resepresents a species. Circles are jittered vertically around the named depths (ticks) to be able to distinguish visually. (Are *R*^2^ and *P*-values needed here?)")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mean_mfac vs. rdi.gr.jpeg"))
```

@@

### ERD not correlated with growth
# Limitations & Future directions:

- not hydrodynamics, 1D to 3D

- Temperate rooting depths shallow?  Shenk and Jackson

- lightening explains 1/3rd of mortality
- forest is realtively stable over 40 years.

-transient depth of water uptake, which is known to vary seasonally (Ellsworth & Sternberg, 2015), 
In fact, associations between soil water access by roots and aboveground drought survival traits such as resistance of xylem to drought‐induced cavitation (Pockman & Sperry, 2000; Pratt et al ., 2007) or phenology (Jackson et al ., 1995; Hasselquist et al ., 2010), are fairly reliable trade‐off functions in plant eco‐physiology.

- Deep ERD species may possibly be associated with locations of greater water availability rather than having deep rootedness. Future studies should employ a distributed hydrological model to separate these factors.

Future studies: high resolution growth data to better constrain the model. 
  -water storage (capacitance) should be considered
  -shifts in root water-uptake by depth
  -deci

@@

```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts}Extreme droughts by census interval and some representative depths"}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q2.5.jpeg"))
```

@@

```{r leaf-cover, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-leaf-cover} Seasonal variation in Leaf Cover Fraction for ", length(unique(sp.leaf_cover.for.model$sp[sp.leaf_cover.for.model$deciduousness != "Evergreen"]))," non-evergreen species at Barro Colorado Island. Leaf Cover Fraction for evergreen species (n = ", length(unique(sp.leaf_cover.for.model$sp[sp.leaf_cover.for.model$deciduousness == "Evergreen"])),") is assumed at 1 through the year.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/leaf.cover_BCI_single_panel_evergreensAt100.jpeg"))
```

>>>>

\
Trait-based plant ecology aims to predict impacts of environmental change on community composition and function using relevant plant traits as currency. To reach this goal it is imperative to establish relationships between plant's environment, traits and demographic outcomes, but this remains to be established for trees' realised water environments and water-access through droughts. Huge efforts have been devoted to linking same-for-all-trees drought indices, above-ground traits and demographic rates. These have resulted in above-ground hydraulic traits such as proxies of hydraulic failure as suggested indicators of mortality risk. A recent meta-analysis found greater drought-induced mortality rates in species with high vulnerability to drought-induced embolism ($\Psi_{88,~Stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~Stem}$), but counter examples also exist (Hoffmann 2011; Paddock 2013), indicating whether plants actually reach critical thresholds of hydraulic failure depends on their realised drought intensity. A rare study that accounted for realised water availability for four tree species in a rare, centennial drought (Texas, USA) found that deep rooted species died less despite being more vulnerable to xylem embolism (high $\Psi_{88,~Stem}$) and having narrower safety-margins ($\Psi_{min} - \Psi_{88,~Stem}$). A species sourcing deep water showed lower hydraulic failure and lower mortality in an extreme drought in an Italian Karst system even though it had narrower hydrulic safety margin and high $\Psi_{88,~Stem}$ than other two species limited to shallow-water. Chitra-Tarak et al. (2018) proposed a general model to link realised water availability to demographic outcomes through droughts at tree community-scale. This study found that species accessing deep-water in a tropical dry forest study system died more in a rare, centennial drought compared to shallow-water species. Authors concluded that this was consistent with expectation of deep water usage as an aquisitive strategy relying on stable and reliable water resource at depth, which exhausted in the multi-annual drought resulting in reduced survival, compared to persistent species associated with shallow-water that was relatively more available via infiltration during this drought. However, species estimates of water-access modelled by this study were not corroborated with rooting/water-sourcing depths, or related to hydraulic traits.

@@

\
We test an overarching hypothesis that interspecific variation in effective rooting depths of trees aligns species on a conservative-acquisitive trade-off axis. Specifically, in this forest (i) dynamics of soil water potential at depth does not cross critical plant thresholds even during multi-annual droughts because of high total annual precipitation to evapotranspiration ratio (not usre if this should be stated as a hypothesis or just an explanation of the results); (ii) trees with deeper effective rooting depths (ERD) show acquisitive hydraulic architecture: higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins; finally, (iii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure. To our knowledge this is the first study to demonstrate a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and mortality outcomes for a tropical forest community.

@@

```{r hydro-calib, dev='png', fig.show='hide', eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap ="\\label{lab-hydro-calib}Model evaluation: Observations (red line) of Volumetric Water Content, Evapotranspiration and stream discharge aginst simulations (blue lines) from 100 best-fit parameter ensembles runs of ELM-FATES. This figure will be reformatted to compose individual figures into one."}
p1 <- cowplot::ggdraw() + cowplot::draw_image("figures/2019-10-14_5000/best-fits/swc_Obs_vs_model_daily_horizontal_with_steph_mean.jpeg", scale = 1)
p2 <- cowplot::ggdraw() + cowplot::draw_image("figures/2019-10-14_5000/best-fits/QRUNOFF_Obs_vs_model_monthly.jpeg", scale = 1)
p3 <- cowplot::ggdraw() +
  cowplot::draw_image("figures/2019-10-14_5000/best-fits/ET_Obs_vs_model_monthly.jpeg", scale = 1)
bottom_row <- cowplot::plot_grid(p2, p3, labels = c('B', 'C'), label_size = 12, ncol = 2)
cowplot::plot_grid(p1, bottom_row, labels = c('A', ''), label_size = 12, ncol = 1, rel_widths = c(2, 1))
```

## Model for Leaf Vulnerability Curves at Community-scale
\
We measured ($K_{x,~Leaf}$ vs. $\Psi_{x,~Leaf}$) on adult trees of `r length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))` common species (see SI x). We fitted exponential, sigmoidal, logistic forms of equations to species-level data as prescribed by [@Guyot:2011] and based on AIC criteria chose exponential form (Fig. S\@ref(fig:obs-plc-curves)), described as

@@

```{r comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-plc-curves} Modelled standardized 'Vulnerability curves' ($K_{x,~Leaf}$ vs. $\\Psi_{x,~Leaf}$) at community-scale (", nrow(data.model.AB$sp)," species) based on species data on $WSG$ & $LMA$ and their relationships with vulnerability curve parameters measured on ", sp.soft.filled$AB.data.model," species. Each line represents a species, color coded by $WSG$.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB.jpeg"))
# knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_LMALAM_AVD_predicted_AB.jpeg"))
```

@@

-- Write about why models were designed this way.

\
\begin{equation}
K_{Leaf,z,n} = Ae^{-B\Psi_{z,n}}
(\#eq:k-soil)
\end{equation}

@@

Growth Data: To minimise effect of light variation among trees and census intervals we selected only large trees (>= 30 cm dbh, diamater at breast height) and species whose maximum height was at least 30 m (hereafter, canopy species). We chose only those trees for which diameter (dbh) record was present across all size censuses (1990, 1995, 2000, 2005, 2010, 2015). We calculated individual tree ($i$) growth rates ($Gr_{i,t}$ (cm.yr^{-1})) for each of the five census intervals ($t$) as the   difference between tree dbh measured in successive censuses, divided by the inter-census interval based on the exact date of measurements. As tree growth rates increase with size, to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census onterval ($Resid_{i,t} = Gr_{i,t} - f(DBH_{i, t})$). The diameter model of growth was fitted based on all trees' data pooled together as there was insufficient species-specific data for large trees (SI ). We then standardised (centered and scaled) an individual tree time series ($Resid_{i,t}$) and obtained species-level (*n* = 32) times-series of medians ($Growth_{t}$) that had data on at least three trees (median = xx).

@@

For each model \@ref(eq:M1)- \@ref(eq:M4) and species combination ($sp$), we obtained goodness-of-fit (*R*^2^) and Pearson's $\rho$ for all soil depths $z$, and for each sp, model combination we chose the $z$ with highest *R*^2^ as the Effective Rooting Depth (ERD); after retaining species with positive $\rho$ and *R*^2^ >= 0.1.

@@

\
Model with VPD and without Leaf Fraction (Eq. \@ref(eq:M2)) had highest goodness-of fit (*R*^2^ = 0.51; *p* = 0.072), followed by followed by model without both VPD and and Leaf Fraction (\@ref(eq:M1); Fig. S\@ref(fig:other-erd-mods)). Models with leaf fraction had only three species with sufficient species-level goodness-of-fit (Eqs. \@ref(eq:M3)-\@ref(eq:M4)) versus seven for models without (Eqs. \@ref(eq:M1)-\@ref(eq:M2)). For rest of the analyses we therefore used species ERD estimates from model with VPD and without Leaf Fraction (Eq. \@ref(eq:M2); Fig. \@ref(fig:erd-iso)).

```{r erd-iso, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-iso}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE; m) versus dry season stable isotopic concentration $\\delta$^2^H (mean $\\pm$ (add error); ‰) for ", length(unique(erd.iso$sp))," canopy species from Barro Colorado Island (data from [@Meinzer:1999]")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_phenology_Meinzer_gr.Psi.VPD.jpeg"))
```

@@

We calculated mortality rates for large trees (>= 10 cm dbh) of species with ERD estimates (all canopy species with maximum height >= 30 m) and average adbunance of at least 20 trees (dropped one species with average abundance of 9 trees). For each of the seven census intervals ($t$) in the 33 year record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (%yr^-1^) for two successives censuses (cen~{1}~ and cen~{2}~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{t}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees in the 50 ha plot of a species present in cen~{1}~ and dead in cen~{2}~, respectively, and t is the duration based on mean dates of cen~{1}~ and cen~{2}~. We calculated species mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{i}\sum_{i=1}^{6} {Mort_{t}}$ and, finally, species mortality anoamalies by intervals as $Mort_{anomaly}$ = $Mort_{t}$ - $Mort_{mean}$. To test whether $ERD$ plays a role in alleviating mortality risk, we regressed species $ERD$ against species ($Mort_{mean}$). To assess whether this may be due to deeper $ERD$ species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{x, ERD}$) were below species $\Psi_{crit}$, defined as $\Psi_{80,~Leaf}$.

@@

\
Species with deeper $ERD$ showed lower mean mortality rates across the 33 year record (1982-2015; *R*^2^ = 0.13, *P* = 0.029, *n* = xx; Figure \@ref(fig:erd-mort)). For five out of seven census intervals, species with shallower $ERD$ had significantly higher mortality anomalies than deeper ERD species; Figure (\@ref(fig:erd-mort-anom)). Analysis of time spent under catastrophic soil drought showed that the proportion of days species $\Psi_{x,~ERD}$ was more negative than species $\Psi_{crit}$ decreased with species ERD (Figure \@ref(fig:erd-time)). Species with ERD deeper than 1 m did not cross species $\Psi_{crit}$ for the entire 25 year period for which soil water potential $\Psi_{x,~z}$ was modelled (1990-2015), as soil depths at and below 1.7 m remained above `r round(psi.1.7.min$median, 2)` [`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`] (median [95% CI]) through all the dryseason and drought periods, which is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).

```{r erd-mort, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort}Mortality rate (mean $\\pm$ 1SE) versus Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) for xx canopy species found on Barro Colorado Island. *R*^2^ and *P*-values are given in panel insets."}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_rate_by rdi.gr.jpeg"))
```

```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{x, z}$ was below $\\Psi_{crit}$, defined as $\\Psi_{80,~Leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for xx canopy species found on Barro Colorado Island. Each circle resepresents a species. Circles are jittered vertically around the named depths (ticks) to be able to distinguish visually. (Are *R*^2^ and *P*-values needed here?)")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mean_mfac vs. rdi.gr.jpeg"))
```

@@

Our analyses revealed that trees with deeper effective rooting depths (ERD) showed acquisitive hydraulic architecture, with higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species. Yet, deeper ERD species showed lower mortality rates over the 33 year record (1982-2015). Analysis of time spent below ($\Psi_{crit}$) in the soil zone of ERD showed that this was likely due to low realised risk of hydraulic failure for deeper ERD species. In this moist forest ($\Psi_{z}$) for depths at and below 1.7 m remained above ($\Psi_{crit}$) of the most sensitive tree species through all the dryseason and drought periods over 1990-2015, period over which ($\Psi_{z}$) was modelled. We thus show, for the first time, that interspecific variation in effective rooting depths of trees aligns species on a conservative-acquisitive trade-off axism, by demonstrating a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a tropical forest community.

## Limitations & Future directions

Our modeling framework had a few limitations. 
- The inverse model for ERD was parametarised on 5-yr growth rates, which vastly increased the degree of freedom to fit the model. Future studies should use high-frequency growth time series such as those obtained from dendrometer bands. To infer growth from such diameter time series, however, would especially require accounting for diameter changes due to hydrostatic flexing [@Chitra-Tarak:2015].

- Future studies should investigate seasonality in water-sourcing depths investigated.

- Although we attempted a coarse estimate of species leaf cover seasonality by combining  leaf-fall data from litter traps and forest-level LAI data for leaf-on periods, more accuarate estimations of leaf-cover seasonality are required, such as those obtained from drone images [@Park: 2019]

-To enable estimation of inverse model for the larger community across size classes would need accounting for changing light environments on growth; we minimised this issue by limiting our ERD inversion to large trees of species with max height >= 30m and thus whose canopy's are most likely to be fully exposed to light.

- Future studies would benefit from modeling the complete hydrodynamics of water potential gradients from soil through plant tissues (roots-stems-leaves) to atmosphere continuum (SPAC) [@Johnson:2017]. However, to parameterise such a model at community-scale would need community-scale data on hard traits such as stem and root vulnerability curves; if not, model development to up-scale data from a few species, for example, via softer traits as we did for leaf vulnerability curves. (Note that In this forest LMA and WSG did not scale with stem vulnerability curves). Water storage (capacitance) should also be considered.

- Deep ERD species may possibly be associated with locations of greater water availability rather than having deep rootedness. A 3-D hydrology model would provide spatially distributed estimates of water dynamics in the whole-soil/rock column and depth to water table to link with individual trees water experience [@Schwantes:2018]. Forest inventory sites would benefit from investing data collection efforts on water availability dynamics by depth and water tables covering spatial heterogeneity so as to better constrain such models.

- Variation in mortality explained by ERD is low. Other causes.... Based on a year-long study, lightening explains ~4.5 % of mortality for trees >=10 cm dbh and ~ 40% for trees >=60 cm dbh.

-Forest is realtively stable over 35 years. Future studies could investigate if ongoing change in the duration, frequency and intensity of droughts entails sufficient exhaustion of deep water availablity so as to cause mortality of deep ERD species.

## Conclusions:

@@

```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts}Extreme droughts by census interval and some representative depths"}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q2.5_depths_in_inverse_model.jpeg"))
```

@@

```{r leaf-cover, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-leaf-cover} Seasonal variation in Leaf Cover Fraction for ", length(unique(sp.leaf_cover.for.model$sp[sp.leaf_cover.for.model$deciduousness != "Evergreen"]))," non-evergreen species at Barro Colorado Island. Leaf Cover Fraction for evergreen species (*n* = ", length(unique(sp.leaf_cover.for.model$sp[sp.leaf_cover.for.model$deciduousness == "Evergreen"])),") is assumed at 1 through the year.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/leaf.cover_BCI_single_panel_evergreensAt100.jpeg"))
```
